---
title: "NOAA Storm Database - Analysis: Severe weather events"
author: "Santiago R"
date: "21 November 2017"
output:
        html_document:
                keep_md: true
---

```{r setup, include=FALSE}
rm(list = ls()) #Delete existing data in global environment
library(knitr)
library(lattice)
library(ggplot2)
library(dplyr)
library(anytime)
opts_chunk$set(echo = TRUE, cache = TRUE) ## Always show answers and avoid repeating unnecessary calculations
```

# Questions:

The data Analysis will address the following questions:


- Across the United States, which types of events (as indicated in the EVTYPE variable) are
most harmful with respect to population health?


- Across the United States, which types of events have the greatest economic consequences?

# Synopsis

Add here the Synopsis! **Pending**

# Data processing

## Reading the Data

``` {r readingData}
# for Reading Files *ZIP
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileUrl, "dataset.zip")
Rawdata <- read.csv("dataset.zip", header = TRUE, sep = ",", check.names = TRUE)

#Check data
str(Rawdata)
``` 

BGN_DATE represent a Date and is being shown as "Factor". TO convert to type: "Date"

``` {r changetodate}
#Make a copy of data to avoid reading again the file
Rawdata2 <- Rawdata
Rawdata2$BGN_DATE <- as.Date(Rawdata2$BGN_DATE, "%Y/%m/%d")
#validate if type of data for "Date" has changed
str(Rawdata$BGN_DATE)

```

One of the Questions is to know which Event Type has been more harmful with respect to
population. 

The first approach is to validate wich Event Type has occurred more times

```{r MostrepeatedEvent}
EventTypeSummary <- table(Rawdata2$EVTYPE)
EventTypeSummary <- data.frame(EventTypeSummary)
EventTypeSummary <- arrange(EventTypeSummary, desc(Freq))
head(EventTypeSummary, 15)
``` 
